---
title: "INSiGHT: A Nonparametric Framework for Multi-Sample Spatially Variable Gene Detection in Spatial Transcriptomics"
author: "Haohao Su"
output:
  html_document:
    toc: true
    toc_depth: 2
    number_sections: true
    code_folding: show
    self_contained: true
params:
  data_file: "../DLPFC/data/DLPFC_151507_151508_151509_151510_combined.RDS"
  use_simulated_data: false
  normalization: "library-size"
  fdr_threshold: 0.05
  sample_adjust_scope: "within-sample"
  plot_gene: ""
  run_fast: false
  export_results: false
  output_dir: "tutorial_output"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = TRUE, message = FALSE,
                     fig.width = 10, fig.height = 5, dpi = 120,
                     collapse = TRUE, comment = "#>", error = FALSE)
```

# Overview

This tutorial (under INSiGHT 1.2.1) identifies spatially variable genes (SVGs) in four DLPFC tissue
sections (151507, 151508, 151509, and 151510). The workflow produces a joint
multi-sample test for each gene and complementary sample-level tests, then
summarizes and visualizes the results. A joint discovery does not establish
that the gene is spatially variable in every section, or that its spatial
expression pattern is identical across sections.

INSiGHT takes one expression matrix and one coordinate matrix per sample.
Samples can have different numbers of spots and do not require spot-to-spot
registration. This does not remove the cross-sample independence assumption of
the joint test: dependence between sections can affect calibration.

# Installation and Setup

Install the updated package in a separate R session before knitting.
```{r install, eval=FALSE, include=FALSE}
if (!require("devtools", quietly = TRUE)){
  install.packages("devtools")
}
devtools::install_github("HaroldSu/INSiGHT")
```

Load **INSiGHT** and other required packages
```{r libraries}
library(INSiGHT)
library(Matrix)
library(ggplot2)

packageVersion("INSiGHT")
```

# Prepare the Input

The input RDS must contain:

| Component | Required structure |
|:--|:--|
| `counts` | List of nonnegative gene-by-spot matrices, preferably sparse, with unique gene and spot names. |
| `spatial.locs` | List of numeric spot-by-two coordinate matrices, with spot names as row names. |
| `samples.id` | Unique sample IDs in the same order as both lists. |

By default, INSiGHT stores library totals before intersecting and filtering genes (`normalization = "library-size"` in `processINSiGHT()`). Therefore, provide the full count matrices rather than preselecting SVGs; otherwise, set `normalization = "none"` when running `processINSiGHT()`.
Gene names identify shared genes across samples. Spot names need to be unique within each sample, but can recur in different samples.

```{r}
dat <- readRDS("./data/DLPFC_151507_151508_151509_151510_combined.RDS")
str(dat)
```

# Standard Analysis

The standard example computes and uses all eigenvalues after constrcuting the $k_s$-nearest-neighbor Gaussian kernel. This can require substantial time and memory for samples with more than 10,000 spots. The optional fast mode below uses a truncated eigendecomposition.

## Create and Process the Object

```{r create-object}
object_INSiGHT <- createINSiGHTobject(data.input = dat$counts,
                     spatial.locs = dat$spatial.locs,
                     sample.names = dat$samples.id)
```

The following settings are explicit so the tutorial does not rely on changing
package defaults:

- Quality control removes spots with total input expression below 10, genes
  detected in fewer than 1% of retained spots in any sample, and mitochondrial
  genes identified by the package's case-insensitive `mt-` name matching.
- `normalization = "library-size"` divides each spot's expression by its stored
  input library total and multiplies by 10,000 before ranking. 
  `"none"` ranks input values directly;
  `"log-library-size"` additionally applies `log1p`, giving equivalent ranks to
  library-size normalization apart from numerical precision.
- `rank.zero = FALSE` assigns zero counts an initial rank of zero and ranks only
  positive values within each gene and sample. Ties receive average ranks by default.
- Bandwidth estimation uses 36 nearest neighbors and quantile 0.3.

Normalization changes the target from absolute count patterns toward relative
expression patterns. It can mitigate spatial library-size effects, but changes
in tissue composition or broad biological expression shifts can also change
library totals. Choose preprocessing according to the scientific question;
normalization is not a guarantee against confounding.

```{r process-standard}
object_INSiGHT <- processINSiGHT(
    object = object_INSiGHT,
    spot.threshold = 10, gene.threshold = 0.01,
    NN4Kernel = FALSE, eigen.less = FALSE,
    kernel.option = "single",
    k.NN4bandwidth = 100, quantile.bandwidth = 0.1,
    rank.tie = "average", rank.zero = FALSE,
    normalization = "library-size"
  )
```

## Joint and Sample-Level Tests

`sample.inference = TRUE` reuses the sample-specific statistics and eigenvalues
to calculate sample-level p-values alongside the joint test. It does not require
reprocessing each sample, and results are returned for all tested genes, not
only genes selected by the joint test.

```{r test-standard}
object_INSiGHT <- testINSiGHT(
    object = object_INSiGHT,
    QuadForm.method = "davies",
    sample.inference = TRUE,
    p.adjust.method = "BY"
  )
```

For the joint analysis, BY adjustment is across all tested genes. With
`sample.adjust.scope = "within-sample"`, sample-level BY adjustment is performed
separately across genes in each sample. Set it to `"all-samples"` to adjust across
all tested gene-sample pairs instead; joint adjustment is unchanged. 

# Inspect the Results

## Joint Discoveries

With the single-kernel setting used here, no across-kernel combination is needed.

`testResults` contains `ID` (gene identifier), `p.values_combined` (unadjusted joint p-value; when `kernel.option = "mxiture"`, unadjusted p-value after Cauchy combination across the results under multiple kernels), and `p.values_adjusted` (adjusted joint p-value). 

```{r joint-results}
knitr::kable(head(object_INSiGHT@testResults, 10), digits = 4,
             caption = "Joint results")
head(object_INSiGHT@testResults)
```

## Complementary Sample-Level Results

Each element of `sampleTestResults` has the same result columns as the joint
table and is named by sample ID. Sample-level significance helps localize
evidence, but a significant result in one sample and a nonsignificant result in
another does not itself demonstrate a between-sample difference.

```{r sample-results}
sample_results <- object_INSiGHT@sampleTestResults
knitr::kable(head(sample_results[[object_INSiGHT@samples[1]]], 10), digits = 4,
             caption = paste("Sample-level results:", object_INSiGHT@samples[1]))
```


# Optional Fast Mode

The fast mode requests the top 10% of eigenvalues. Approximation can affect calibration and power.

```{r fast-mode}
object_fast <- processINSiGHT(
  object = object_INSiGHT,
  spot.threshold = 10, gene.threshold = 0.01,
  NN4Kernel = TRUE, k.NN4Kernel = 100,
  eigen.less = TRUE, eigen.prop = 0.1,
  rank.tie = "average", rank.zero = FALSE,
  normalization = "library-size"
)
object_fast <- testINSiGHT(
  object_fast, sample.inference = TRUE
)
  
knitr::kable(head(object_fast@testResults, 10), digits = 4,
             caption = "Joint results (fast mode)")
```
